% SPDX-License-Identifier: AGPL-3.0-or-later
% SPDX-FileCopyrightText: University of Washington <https://www.washington.edu>
% SPDX-FileContributor: 2014-22 Bradley M. Bell
% ----------------------------------------------------------------------------
% Beamer template file.

\documentclass{beamer}

% Choose theme; e.g., default, Wasaw
\usetheme{default}

% To avoid font warnings caused by Beamer
\usepackage{lmodern}

% For web links
\usepackage{hyperref}

% use lable instead of default icon in bibliography
\setbeamertemplate{bibliography item}{\insertbiblabel}

% Binary file not included for this talk
% \pgfdeclareimage[height=0.5cm]{ihme-apl-logo}{ihme-apl-logo}
% \logo{\pgfuseimage{ihme-apl-logo}}

% Common macros
\newcommand{\Green}[1]{\textcolor{green}{#1}}
\newcommand{\Blue}[1]{\textcolor{blue}{#1}}
\newcommand{\B}[1]{{\bf #1}}
\newcommand{\R}[1]{{\rm #1}}
\newcommand{\Section}[1]{
	\section{#1}
	\begin{frame}
	\begin{center}
	\Blue{ \Large{#1} }
	\end{center}
	\end{frame}
}
%
\title[cppad\_mixed]{
The cppad\_mixed Capture Example and Speed Test}
\author{Bradley M. Bell}
\institute{
	Applied Physics Laboratory, \\
	Health Metrics and Evaluation, \\
	University of Washington, \\
	{\tt bradbell@uw.edu}
}
\date[2016-06-(20-24)]
{ADMB and TMB Developer's Workshop \\ June, 2016}
%
% This example definition is intended to be replaced by a renewcommand that
% defines a specfic \TabbedItem just before it is used.
\newcommand{\TabbedItem}[1]{
\item
\begin{tabbing}
12345\=67890\=12345\=67890\=12345\=\kill
#1
\end{tabbing}
}

\begin{document}

\begin{frame}
	\titlepage
\end{frame}

\begin{frame}{Contents}
\tableofcontents[pausesections]
\end{frame}


% ===========================================================================
\Section{Introduction}
% --------------------------------------------------------------------------

\begin{frame}{CppAD Mixed}
C++ Laplace approximation for mixed effect models
\cite{BellCppADMixed}.
\pause

\begin{itemize}

\item
Implements the method in TMB paper \cite{Kristensen2016}.
\pause

\item
Can be Back-end for TMB in ADMB, Stan, python, ... \ldots
\pause

\item
Starting point and proposal sampling for MCMC methods.
\pause

\item
Separate fixed and random likelihoods.
\pause

\item
Gaussian or Laplace priors (Lasso) on fixed effects.
\pause

\item
Constraints on the fixed effects, on random effects.
\pause

\item
No user defined C++ template functions.
\pause

\item
CppAD add-on, traps CppAD errors, recovers from errors.
\pause

\item
Easy debugging and asserts in complex modeling situations.
\pause

\end{itemize}
\end{frame}

% --------------------------------------------------------------------------

\begin{frame}{Install}
Looking for package managers for particular systems:
\pause

\begin{itemize}

\item
bash shell and web get program wget.
\pause

\item
Compilers: C++ and Fortran.
\pause

\item
gsl: The GNU scientific library.
\pause

\item
cmake: cross platform configuration and installer.
\pause

\item
package-config: information about previous installs.
\pause

\item
Set the install prefix and choose debug / relase for:
\pause

\begin{itemize}
\item
bin/install\_eigen.sh
\pause
\item
bin/install\_ipopt.sh
\pause
\item
bin/install\_suitesparse.sh
\pause
\item
bin/install\_cppad.sh
\pause
\end{itemize}

\item
bin/run\_cmake.sh: edit the run.
\pause

\item
make check, make speed, make install
\pause

\item
bin/debian\_install.sh: puts it all together for a Debian system
\pause

\end{itemize}
\end{frame}
% --------------------------------------------------------------------------

\begin{frame}{capture\_xam}
This is a \texttt{cppad\_mixed} example; see  \cite{Royle2004}.
\pause

\begin{itemize}

\item
Tests both optimization and posterior sampling.
\pause

\item
Reports timing test results.
\pause

\item
Reports correctness test results.
\pause

\item
Reports memory usage results.
\pause

\item
Separate control of data set size.
\pause

\item
Separate control of number of random effects.
\pause

\item
Separate control of computational load.
\pause

\item
Compare Newton and quasi-Newton methods.
\pause

\item
Interesting numerical AD method used to avoid overflow.
\pause

\item
Optional constraints.

\end{itemize}
\end{frame}

% ===========================================================================
\Section{Data Model}
% --------------------------------------------------------------------------

\begin{frame}{Data Given Population Size and Capture Probability}
\begin{itemize}

\item
$y_{i,t}$ is the number of captures at location $i$ and time $t$.
\pause

\item
$N_i$ is the size of the population at location $i$.
\pause

\item
$q_t$ is the probability of capture at time $t$.
\pause

\item
The conditional probability of $y_{i,t}$ given $N_i$ and $q_t$ is
\[
\B{p} ( y_{i,t} | N_i , q_t )
=
\left( \begin{array}{c} N_i \\ y_{i,t} \end{array} \right)
q_t^{y(i,t)} \left( 1 - q_t\right)^{y(i,t)}
\]

\end{itemize}
\end{frame}
% --------------------------------------------------------------------------

\begin{frame}{Population Size Given Fixed Effects}
\begin{itemize}

\item
$\theta_0$ is the mean of $N_i$ .
\pause

\item
$\theta$ is the vector of fixed effects.
\pause

\item
We use a Poisson distribution to model the probability of $N_i$
given $\theta$
\[
\B{p} ( N_i | \theta  )
=
\theta_0^{N(i)} \frac{ \exp[ - \theta_0 ] }{ N_i ! }
\]

\end{itemize}
\end{frame}
% --------------------------------------------------------------------------

\begin{frame}{Capture Probability Given Fixed and Random Effects}
\begin{itemize}

\item
$\theta_1$ is the mean of the logit of $q_t$.
\pause

\item
$u_t$ is the random effect at time $t$.
\pause

\item
$u$ is the vector of random effects.
\pause

\item
The logit of $q_t$ is
\[
	\log [ q_t / ( 1 - q_t ) ] = u_t + \theta_1
\]
\pause

\item
The probability of capture at time $t$ given $\theta$ and $u$ is
\[
	q_t ( \theta , u )
	=
	\B{p} ( q_t | \theta , u )
	=
	[ 1 + \exp( - u_t - \theta_1 ) ]^{-1}
\]


\end{itemize}
\end{frame}
% --------------------------------------------------------------------------

\begin{frame}{Data Given Fixed and Random Effects}
\begin{itemize}

\item
$y_i$ vector of data at location $i$ and all times.
\pause

\item
$T$ is the number of times.
\pause

\item
The probability of $y_i$ given $N_i$, $\theta$, and $u$ is
\[
\B{p}( y_i | N_i, \theta , u )
=
\prod_{t=0}^{T-1}
\left( \begin{array}{c} {N(i)} \\ y_{i,t} \end{array} \right)
	q_t ( \theta , u)^{y(i,t)}
	\left( 1 - q_t( \theta , u) \right)^{y(i,t)}
\]
\pause

\item
$K$ is maximum population in truncation of infinite summation.
\pause

\item
Probability of $y_i$ given $\theta$, $u$, is modeled using the sum
\[
\B{p}( y_i | \theta , u )
=
\sum_{k=0}^K \B{p}( y_i | N_i=k, \theta , u ) \B{p}( N_i=k | \theta )
\]
\pause

\end{itemize}
\end{frame}
% --------------------------------------------------------------------------
%
\begin{frame}{Data Given Fixed and Random Effects}
\begin{itemize}

\item
$y$ vector of data at all locations and all times.
\pause
%
\item
$R$ is the number of locations.
\pause
%
\item
The probability of $y$ given $\theta$, and $u$ is
\[
\B{p}( y | \theta , u )
=
\prod_{i=0}^{R-1} \B{p}( y_i | \theta , u )
=
\]
\pause
\[
\prod_{i=0}^{R-1}
\left[
\sum_{k=0}^{K-1}
\theta_0^k \frac{ \exp[ - \theta_0 ] }{ k ! }
\prod_{t=0}^{T-1}
\left( \begin{array}{c} {k} \\ y_{i,t} \end{array} \right)
	q_t ( \theta , u)^{y(i,t)}
	\left( 1 - q_t( \theta , u) \right)^{y(i,t)}
\right]
\]

\end{itemize}
\end{frame}
% ===========================================================================
\Section{Program Input}
% --------------------------------------------------------------------------

\begin{frame}{Random Likelihood Function}
\begin{itemize}

\item
$\theta_2$ is the standard deviation of the logit of $q_t$.
\pause

\item
We use a normal distribution for $u$ given $\theta$,
\[
\B{p} ( u | \theta )
=
\prod_{t=0}^{T-1}
	\frac{1}{ \theta_2 \sqrt{ 2 \pi } }
		\exp \left[ - \frac{1}{2} \frac{ u_t^2 }{ \theta_2^2 } \right]
\]
\pause

\item
The user defined random likelihood for this example is
\[
f( \theta , u )
=
- \log [ \B{p} ( y | \theta , u) \B{p} ( u | \theta ) ]
\]

\end{itemize}
\end{frame}
% --------------------------------------------------------------------------
\begin{frame}{Other User Defined Functions}
\begin{itemize}

\item
There is no prior for the fixed effects; i.e.,
$\B{p}( \theta )$.
\pause

\item
There is no data that only depends on fixed effects; i.e.,
$\B{p}(z| \theta )$.
\pause

\item
There is no fixed constraint functions; i.e,
$c( \theta )$.
\pause

\item
\[
\hat{u} ( \theta ) = \R{argmin} \; f( \theta , u ) \; \R{w.r.t.} \; u
\]
\pause

\item
$A$ is the random constraint matrix.

\item
The random constraint function is
\[
0
=
A \hat{u} ( \theta )
=
\hat{u}_1 ( \theta ) + \cdots + \hat{u}_{T-1} ( \theta )
\]

\end{itemize}
\end{frame}
% --------------------------------------------------------------------------
\begin{frame}{Command Line Arguments}
\fontsize{10}{11}\selectfont
\begin{itemize}

\item
random\_seed
\hspace{4em} non-negative integer
\hspace{0em} 0 , match previous
\pause

\item
number\_fixed\_samples
\hspace{0.2em} positive integer
\hspace{2.1em} 1000
\pause

\item
number\_locations
\hspace{2.0em} positive integer , $R$
\hspace{0.6em} 50
\pause

\item
number\_times
\hspace{3.5em} positive integer , $T$
\hspace{0.5em} 10
\pause

\item
max\_population
\hspace{2.7em} positive integer , $K$
\hspace{0.5em} 50
\pause

\item
mean\_population
\hspace{2.2em} positive real , $\theta_0$
\hspace{1.8em} 5.0
\pause

\item
mean\_logit\_probability
\hspace{0em} real , $\theta_1$
\hspace{5.1em} -0.5
\pause

\item
std\_logit\_probability
\hspace{1em} positive real , $\theta_2$
\hspace{1.9em} 0.5
\pause

\item
quasi\_fixed
\hspace{4.8em} true or false
\hspace{3.8em} true
\pause

\item
random\_constraint
\hspace{1.6em} true or false
\hspace{3.8em} false , true
\pause

\item
trace\_optimize\_fixed
\hspace{1em} true or false
\hspace{3.8em} false
\pause

\end{itemize}
\end{frame}
% ===========================================================================
\Section{Program Output}
% --------------------------------------------------------------------------
\begin{frame}{Seed, Memory, Timing Results}
\begin{itemize}

\item
actual\_seed
\hspace{5.9em} 1466520673 , 1466520673
\pause

\item
initialize\_bytes
\hspace{4.6em} 27,728,078 , 27,728,078
\pause

\item
initialize\_seconds
\hspace{3.6em} 0.244 , 0.259
\pause

\item
optimize\_fixed\_seconds
\hspace{1.3em} 5.49 , 2.96
\pause

\item
optimize\_random\_seconds
\hspace{0.05em} 0.077 , 0.101
\pause

\item
information\_mat\_seconds
\hspace{0.25em} 1.24 , 1.21
\pause

\item
sample\_fixed\_seconds
\hspace{1.9em} 0.024 , 0.041
\pause

\end{itemize}
\end{frame}

% --------------------------------------------------------------------------
\begin{frame}{Estimate Results}
\begin{itemize}

\item
sum\_random\_effects
\hspace{4.9em} -0.0698 , 1.33227e-15
\pause

\item
mean\_population\_estimate
\hspace{2.5em} 5.84 , 5.35
\pause

\item
mean\_logit\_probability\_estimate
\hspace{0.0em} -0.607 , -0.494
\pause

\item
std\_logit\_probability\_estimate
\hspace{1.3em} 0.583 , 0.667
\pause

\item
mean\_population\_std
\hspace{4.7em} 0.579 , 0.492
\pause

\item
mean\_logit\_probability\_std
\hspace{2.5em} 0.23 , 0.092
\pause

\item
std\_logit\_probability\_std
\hspace{3.5em} 0.142 , 0.161
\pause

\item
mean\_population\_ratio
\hspace{3.7em} 1.45 , 0.702
\pause

\item
mean\_logit\_probability\_ratio
\hspace{1.5em} -0.464 , 0.07
\pause

\item
std\_logit\_probability\_ratio
\hspace{2.5em} 0.584 , 1.04
\pause


\end{itemize}
\end{frame}
% --------------------------------------------------------------------------
\begin{frame}{Averages for Twenty Runs Each}
\begin{itemize}

\item
optimize\_fixed\_seconds\_avg
\hspace{2.3em} 5.59 , 2.71
\pause

\item
mean\_population\_error\_avg
\hspace{2.35em} 0.41 , 0.37
\pause

\item
mean\_population\_std\_avg
\hspace{3.05em} 0.52 , 0.49
\pause

\item
mean\_logit\_probability\_error\_avg
\hspace{0.2em} 0.18 , 0.10
\pause

\item
mean\_logit\_probability\_std\_avg
\hspace{0.9em} 0.21 , 0.09
\pause

\item
std\_logit\_probability\_error\_avg
\hspace{1.2em} 0.13 , 0.13
\pause

\item
std\_logit\_probability\_std\_avg
\hspace{1.85em} 0.12 , 0.12
\pause


\end{itemize}
\end{frame}

% ===========================================================================
\Section{Bibliography}
%
\bibliography{capture_xam}{}
\bibliographystyle{alpha}


\end{document}
